\section{04/05/2024}

\subsection{Semi-direct products}

We first start with an exercise of \textbf{direct products}.

\begin{exercise}
\index{Direct product of groups}
\label{xca:direct_product}
Let $G$ be a group and $H$ and $K$ be normal subgroups of $G$.
If $G=HK$ and $H\cap K=\{1\}$, then $G\simeq H\times K$.
\end{exercise}

% \begin{proof}
% Sea $f\colon G\to H\times K$, $f(g)=(h,k)$, donde $h\in H$ y $k\in K$ son únicos tales que $g=hk$. Esto tiene sentido pues si $g\in G$ then $g=hk$ para algún $h\in H$ y $k\in K$; si además $g=h_1h_1$ para $h_1\in H$ y $k_1\in K$, then, como $hk=h_1k_1$, se tiene que $h_1^{-1}h=k_1k^{-1}\in H\cap K=\{1\}$ y luego $h=h_1$ y $k=k_1$.

% Veamos que si $g=hk$ y $g_1=h_1k_1$ para $h,h_1\in H$ y $k,k_1\in K$, then $kh_1=h_1k$. En efecto, $[k,h_1]=kh_1k^{-1}h_1^{-1}\in H\cap K=\{1\}$ pues la normalidad de $H$ y $K$ implican que $ kh_1k^{-1}\in H$ y $h_1k^{-1}h_1^{-1}\in K$.

% La observación anterior nos permite demostrar que $f$ es un morfismo de grupos. Si $g=hk$ y $g_1=h_1k_1$ con $h,h_1\in H$ y $k,k_1\in K$, then, como $f(g)=(h,k)$ y $f(g_1)=(h_1,k_1)$, tenemos que
% \[
% f(gg_1)=f((hk)(h_1k_1))=f(h(kh_1)k_1)=f((hh_1)(kk_1))=(hh_1,kk_1).
% \]
% Queda como ejercicio demostrar que $f$ es biyectiva.
% \end{proof}

\begin{exercise}
Let $A$ be a normal subgroup of $H$, and $B$ be a normal subgroup of $K$. 
Prove that $A\times B$ is a normal subgroup of
$H\times K$ and 
\[
\frac{H\times K}{A\times B}\simeq(H/A)\times(K/B).
\]
\end{exercise}

\index{Exact factorization of groups}
We say that a group $G$ admits an exact factorization through 
the subgroups $H$ and $K$ if $G=HK$ and 
$H\cap K=\{1\}$. By Exercise \ref{xca:direct_product}, 
if $G$ admits an exact factorization through two normal subgroups, then 
it is isomorphic to the direct product of these subgroups. 

\begin{exercise}
    Let $G$ be a group that admits an exact factorization through
    the subgroups $H$ and $K$. Prove that every $g\in G$ can be written 
    uniquely as $g=hk$ for some $h\in H$ and $k\in K$. 
\end{exercise}

\begin{example}
Let $G=\Sym_3$, $H=\langle (123)\rangle\unlhd G$ and $K=\langle (12)\rangle$. Since
$K$ is not normal in $G$, we cannot apply Exercise \ref{xca:direct_product}. 
We do have $G=HK$ and $H\cap K=\{\id\}$, but $H\times K\simeq\Z/3\times\Z/2\not\simeq\Sym_3$, 
as $\Z/3\times\Z/2$ is abelian and $\Sym_3$ is not. 
\end{example}


\begin{definition}
\index{Complement}
Let $G$ be a group, $K$ a normal subgroup of $G$, and $Q$ a subgroup of $G$. We say
that $Q$ \textbf{complements} $K$ in $G$ if $K\cap Q=\{1\}$ y $G=KQ$.
\end{definition}

\begin{example}
Let $G=\Sym_3$ and $K=\langle (123)\rangle\unlhd G$. The subgroups
$\langle (12)\rangle$, $\langle
(13)\rangle$ and $\langle (23)\rangle$ complement $K$ in $G$.
\end{example}

The previous example shows that complements are not unique. However, 
complements are unique under isomorphism, as 
\[
G/K= KQ/K\simeq Q/K\cap Q=Q/\{1\}\simeq Q.
\]

\begin{definition}
\index{Semi-direct product}
We say that a group $G$ is a \textbf{semi-direct product} of $Q$ and $K$ if $K$ 
is normal in $G$ and 
$K$ admits a complement in $G$ isomorphic to $Q$. Notation: $G=K\rtimes Q$.
\end{definition}

Let us discuss some characterizations:

\begin{proposition}
Let $K$ be a normal subgroup of $G$. The following statements are equivalent:
\begin{enumerate}
\item $K$ admits a complement in $G$.
\item There exists a subgroup $Q$ of $G$ such that each $g\in G$ can be written uniquely 
as $g=xy$ for some 
$x\in K$ and $y\in Q$.
\item There is a group homomorphism $s\colon G/K\to G$ such that $\pi\circ s=\id_{G/K}$, where $\pi\colon G\to G/K
$, $g\mapsto Kg$, is the canonical homomorphism.
\item There exists a group homomorphism $\rho\colon G\to G$ such that $\ker\rho=K$ and the restriction $\rho|_{\rho(G)}$ equals the identity. 
\end{enumerate}
\end{proposition}

\begin{proof}
We first prove that $(1)\implies(2)$. If $Q$ complements $K$, then $G=KQ$ and
$K\cap Q=\{1\}$. In particular, if $g\in G$, then $g=xy$ para $x\in K$ e $y\in Q$. To show that
the decomposition is unique, suppose that 
$g=x_1y_1$ with $x_1\in K$ and $y_1\in Q$. Then $x_1^{-1}x=yy_1^{-1}\in K\cap Q=\{1\}$ 
and hence $x=x_1$ and $y=y_1$.

It is an exercise to show that $(2)\implies(1)$.

We now prove that $(2)\implies(3)$. Let $s\colon G/K\to G$, $s(Kg)=y$ if 
$g=xy$ with $x\in K$ and $y\in Q$. (Note that here we use right cosets, as it is more convenient.)
Let us check that $s$ is well-defined. 
For that purpose, we must show that $Kg=Kg_1$ implies $s(Kg)=s(Kg_1)$. If $g=xy$ 
and $g_1=x_1y_1$ with $x,x_1\in K$ and $y,y_1\in Q$, then, since $Kg=Kg_1$, 
$xyy_1^{-1}x_1^{-1}=gg_1^{-1}\in K$, that is $yy_1^{-1}\in x^{-1}Kx_1=K$
 because $x,x_1\in K$. Hence $yy_1^{-1}\in K\cap Q=\{1\}$ and thus $y=y_1$. 
 We now show that $\pi\circ
 s=\id_{G/K}$. If $g=xy$ with $x\in K$ e $y\in Q$, then
$(\pi\circ s)(Kg)=\pi(y)=Ky=Kxy=Kg$. It is left to the reader to prove that 
$s$ is a group homomorphism. 

We now prove that $(3)\implies(4)$. Let $\rho=s\circ\pi$. Then $\rho$ is a group homomorphism (because it is the composition of homomorphisms). We compute: 
\[
\rho(\rho(g))=\rho( (s\circ\pi)(g))=\rho(s(Kg))=((s\circ\pi)\circ s)(Kg)=s(Kg)=\rho(g).
\]
We now compute $\ker\rho$. If $g\in\ker\rho$, then $s(\pi(g))=\rho(g)=1$. Thus 
\[
\pi(g)=\pi(s(\pi(g)))=\pi(1)=1_{G/K},
\]
that is $g\in\ker\pi=K$. Conversely, if $x\in K$, then
\[
\rho(x)=\rho(s(Kx))=\rho(s(K))=\rho(1)=1
\]
and hence $x\in\ker\rho$.

Finally, we prove that $(4)\implies(1)$. We claim that $Q=\rho(G)$ complements $K$ in
 $G$. We first show that $K\cap Q=\{1\}$: if $x\in K\cap Q$, then $x=\rho(g)$ for some $g\in G$. Moreover, 
\[
1=\rho(x)=\rho(\rho(g))=\rho(g).
\]
Hence $g\in\ker\rho=K$ and $x=1$. We now prove that $G=KQ$. For the inclusion $G\subseteq KQ$, 
\[
g=(g\rho(g^{-1}))\rho(g)
\]
and $g\rho(g^{-1})\in K=\ker\rho$, as $\rho(g\rho(g^{-1}))=  \rho(g)\rho(g^{-1})=1$.
\end{proof}

\begin{example}
$\Sym_n=\Alt_n\rtimes\Z/2$, as $Q=\langle (12)\rangle\simeq\Z/2$ complements 
$\Alt_n$ in $\Sym_n$.
\end{example}

For a group $G$, 
the set 
\[
\Aut(G)=\{f\colon G\to G:f\text{ bijective}\}
\]
is a group with the composition of maps. It is called 
the \textbf{automorphism group} of $G$. 
Examples of automorphism groups are the identity map and 
conjugations homomorphisms. 

\begin{example}
$\Aut(\Z)\simeq\Z/2$, as $\Aut(\Z)=\{\id,-\id\}$.
\end{example}

\begin{example}
Let $G$ be a group and $g\in G$. The conjugation 
map $\gamma_g\colon G\to G$, $x\mapsto gxg^{-1}$,
is an automorphism of $G$, as 
\[
\gamma_g(xy)=g(xy)g^{-1}=(gxg^{-1})(gyg^{-1})=\gamma_g(x)\gamma_g(y).
\]
Moreover, $\gamma\colon G\to\Aut(G)$, $g\mapsto\gamma_g$, is a group
homomorphism:
\[
\gamma_{gh}(x)=(gh)x(gh)^{-1}=g(\gamma_h(x))g^{-1}=\gamma_g(\gamma_h(x))=(\gamma_g\circ\gamma_h)(x).
\]

\index{Inner automorphisms}
The group of \textbf{inner automorphisms} of $G$ is the 
group 
$\Inn(G)=\gamma(G)$. Note that $\ker\gamma=Z(G)$, as
if $g\in G$ is such that $\gamma_g=\id$, then 
\[
\gamma_g(x)=gxg^{-1}=x
\]
for all $x\in G$. By the first isomorphism theorem, 
\[
G/Z(G)\simeq \gamma(G)=\Inn(G).
\]
\end{example}

\begin{exercise}
\label{xca:aut(S3)}
    Prove that $\Aut(\Sym_3)\simeq\Sym_3$. 
\end{exercise}

\begin{exercise}
    Let $G$ be a group. Prove that $\Inn(G)$ is a normal subgroup of $\Aut(G)$. 
\end{exercise} 

For a group $G$, the quotient $\Aut(G)/\Inn(G)$ is called 
the group of \textbf{outer automorphisms} of $G$. 
Note that 
\[
\Inn(G)\text{ is cyclic}\Longleftrightarrow
|\Inn(G)|=1\Longleftrightarrow
G\text{ is abelian.}
\]

\begin{exercise}
Let $G$ be a group. 
Prove that if $\Aut(G)$ is cyclic, then 
$G$ is abelian. 
\end{exercise}

\begin{exercise}
    Prove that if $G$ is finite, then $\Aut(G)$ is finite.
\end{exercise}

% \begin{exercise}
% \label{xca:autgeq2}
%         Si $G$ es un grupo tal que $|G|\geq3$, entonces $|\Aut(G)|\geq2$.
% \end{exercise}


% \begin{exercise}
% \label{xca:aut_impar}
%         No existe un grupo finito cuyo grupo de automorfismos sea no trivial cíclico y de orden impar.
% \end{exercise}

% \begin{exercise}
% \label{xca:p2dividesAut}
%         Sea $p$ un número primo.
%         Si $G$ es un $p$-grupo no abeliano, entonces $p^2$ divide a $|\Aut(G)|$.
% \end{exercise}

% \begin{exercise}
%         \label{xca:Aut(H)}
%         Si $G=H\times K$, entonces $\Aut(H)$ es isomorfo a un subgrupo de $\Aut(G)$.
% \end{exercise}

% \begin{exercise}
% \label{xca:aut_trivial_center}
%         Si $G$ tiene centro trivial, entonces $\Aut(G)$ también.
% \end{exercise}

The following exercise constructs semi-direct products. 

\begin{exercise}
\label{xca:semi-direct}
Let $K$ and $Q$ be groups and $\theta\colon Q\to\Aut(K)$, $x\mapsto\theta_x$, a group
homomorphism. Prove that $K\times Q$ 
with 
\[
(a,x)(b,y)=(a\theta_x(b),xy)
\]
is a group. This group will be written as $K\rtimes_\theta Q$.
\end{exercise}

% \begin{proof}[Bosquejo de la demostración]
% Dejamos como ejercicio verificar que la operación es asociativa. Hay que verificar además que el elem
% ento neutro de $K\rtimes_\theta Q$ será $(1,1)$ y
% que el inverso de $(a,x)\in K\rtimes_\theta Q$ será $(\theta_{x^-1}(a^{-1}),x^{-1})$.
% \end{proof}

The group of Exercise \ref{xca:semi-direct} is the
semi-direct product of the subgroups 
\begin{align*}
K\times\{1\}=\{(a,1):a\in K\}\simeq K,&&
\{1\}\times Q=\{(1,x):x\in Q\}\simeq Q
\end{align*}
of $K\rtimes_\theta Q$. Note that $K\times\{1\}$ is normal in $K\rtimes_\theta Q$. 
We can identity $K\rtimes\{1\}$ with $K$ 
and $\{1\}\rtimes Q$ with $Q$. Thus we can write 
\[
\theta_x(a)=xax^{-1}
\]
for all $x\in Q$ and $a\in K$.


\begin{exercise}
Prove that if $G$ is a semi-direct product of the normal subgroup 
$K$ with the subgroup $Q$, there exists a group homomorphism 
$\theta\colon Q\to\Aut(K)$
such that $G\simeq K\rtimes_\theta Q$.
\end{exercise}

% \begin{proof}[Bosquejo de la demostración]
% Para $x\in Q$ sea $\theta_x\colon K\to K$, $\theta_x(a)=xax^{-1}$. Ya vimos que $\theta_x\in\Aut(K)$
% y que $Q\to\Aut(K)$, $x\mapsto\theta_x$ es un morfismo de grupos. Queda verificar que
% la función $K\rtimes_\theta Q\to G$, $(a,x)\mapsto ax$, es un morfismo biyectivo de grupos.
% \end{proof}

Let us discuss some examples. 

\begin{example}
Let $N\simeq \Z/n$ and $H\simeq\Z/2=\{0,1\}$. The map $\theta\colon H\to\Aut(N)$, 
$1\mapsto (x\mapsto x^{-1})$, is a group homomorphism. Let $G=N\rtimes_\theta H$.
Then $G\simeq\D_n$, the dihedral group of order $2n$.
Recall that 
\[
\D_n=\langle r,s:r^n=s^2=1, srs^{-1}=r^{-1}\rangle.
\]
Assume that $N=\langle x\rangle$ and $H=\langle y\rangle$. Then $|(x,1)|=n$ and $|(1,y)|=2$. 
Moreover,
\begin{align*}
(1,y)(x,1)(1,y)^{-1} &= (\varphi_y(x),y)(1,y)\\
&=(\varphi_y(x),y^2)\\
&=(\varphi_y(x),1)\\
&=(x^{-1},1)\\
&=(x,1)^{-1}.
\end{align*}
If $u=(x,1)$ y $v=(1,y)$, then $u^n=v^2=(1,1)$ and $vuv^{-1}=u^{-1}$. Thus there exists 
a surjective group homomorphism 
$\D_n\to G$ (because $G$ is generated by $u$ and $v$). Moreover, $|G|=|N||H|=2n$. Hence 
$G$ has order $2n$ and therefore $G\simeq\D_n$.
\end{example}

\begin{example}
Let $K=\{\id,(12)(34),(13)(24),(14)(23)\}$. Then $K$ is normal in $\Alt_4$. 
Let $H=\langle (123)\rangle\simeq\Z/3$. Since $K\cap H$ is a subgroup of $H$ and $K$ and, moreover, 
$K$ and $H$ have coprime orders, $H\cap K=\{\id\}$. Hence $\Alt_4=K\rtimes H$.
\end{example}

\begin{example}
Let 
\[
K=\{\id,(12)(34),(13)(24),(14)(23)\},\quad  
H=\{\sigma\in\Sym_4:\sigma(4)=4\}.
\]
Note that $H$ is a subgroup of 
$\Sym_4$ isomorphic to $\Sym_3$. Then $H\cap K=\{\id\}$ and hence 
$\Sym_4=K\rtimes H$.
\end{example}

Let $n\geq5$. 
Using the fact that $\Alt_n$ is a simple group,  
one proves that $\Alt_n$ cannot be written as a semi-direct product of proper subgroups. 

\begin{example}
Let $K=\Z/3$ and $Q=\Z/4$. Since $\Hom(Q,\Aut(K))=\{1,\tau\}$, where 
\[
\tau\colon\Z/4\to\Aut(\Z/3)=\{\id,\rho\}\simeq\Z/2,\quad 1\mapsto\rho,
\]
the semi-direct product $T=K\rtimes_\tau Q$ is a non-abelian group of order 12. Moreover,
$T\not\simeq\Alt
_4$ as $|(2,2)|=6$ and $\Alt_4$ has no elements of order six.
\end{example}



% todo: El grupo Aff(R) es un producto semidirecto
% todo: Necesitamos más ejemplos de producto semidirecto
\subsection{Actions}

\begin{definition}
\index{Action}
    Let $G$ be a group and $X$ a set. 
    A (left) \textbf{action} of $G$ on $X$ is a map 
    $G\times X\to X$, $(g,x)\mapsto g\cdot x$, such that 
    \begin{enumerate}
        \item $1\cdot x=x$ for all $x\in X$, y
        \item $g\cdot (h\cdot x)=(gh)\cdot x$ for all $g,h\ n G$ and $x\in X$.
    \end{enumerate}
\end{definition}

If a group $G$ acts on a set $X$, we also say that
$X$ is a $G$-set.

\begin{example}
\index{Action!trivial}
Every group $G$ acts on $G$ trivially: $g\cdot h=h$ for all $g,h\in G$.
\end{example}

\begin{example}
\index{Action!left multiplication}
Every group $G$ acts on $G$ by left multiplication, that is
$g\cdot h=gh$ for all $g,h\in G$.
\end{example}

\begin{example}
\index{Action!conjugation}
Every group $G$ acts on $G$ by conjugation, that is
$g\cdot h=hg^{-1}$ for all $g,h\in G$. More generally, 
if $N$ is a normal subgroup of $G$, then $G$ acts on
$N$ by conjugation: $g\cdot x=gx
g^{-1}$ for all $g\in G$ y $x\in N$. 
\end{example}

\begin{example}
\index{Action!on left cosets}
Let $G$ be a group $H$ be a subgroup of $G$. Then $G$ 
acts on the set of left cosets $G/H$ by left multiplication, that 
is $g\cdot (xH)=(gx)H$ for all $g,x\in G$.
\end{example}

There is a bijective correspondence between 
left actions of a group $G$ on a set $X$ and
group homomorphisms 
$\rho\colon G\to\Sym_X$. The correspondence is given by
the formula 
\[
\rho(g)(x)=g\cdot x,\quad g\in G,x\in X.
\]
We will write $\rho_g=\rho(g)$.

As an example, if $G\times X\to X$, $(g,x)\mapsto x$, is 
an action of $G$ on $X$, then
each $\rho_g\colon X\to X$ is a bijective map with inverse 
$(\rho_g)^{-1}=\rho_{g^{-1}}$. Moreover, 
 $\rho$ is a group homomorphism, as 
\[
\rho(gh)(x)=(gh)\cdot x=g\cdot (h\cdot x)=\rho_g(h\cdot x)=\rho_g(\rho_h(x))
\]
for all $g,h\in G$ and $x\in X$.

\begin{example}
Let $G=\Sym_3$ and 
\[
H=\langle (123)\rangle=\{\id,(123),(132)\}.
\]
Let $G$ act on the set $X=G/H=\{H,(12)H\}$ of left cosets of $H$ 
by left multiplication. Write 
$x_1=H$ and $x_2=(12)H$. Then
\begin{align*}
&(12)\cdot x_1=x_2,
&&(12)\cdot x_2=x_1,
&&(123)\cdot x_1=x_1,
&&(123)\cdot x_2=x_2.
\end{align*}
Since $G=\langle (12),(123)\rangle$, one has the group 
homomorphism 
$\rho\colon G\to\Sym_{X}\simeq\Sym_2$ given by
$(12)\mapsto (12)$, $(123)\mapsto\id$.
\end{example}

\begin{example}
As before, let $G=\Sym_3$ and $H=\langle (12)\rangle=\{\id,(12)\}$. Let $G$ act on the set $X=G/H=\{H,(123
)H,(132)H\}$ of left cosets of $H$ by left multiplication. Write 
$x_1=H$, $x_2=(123)H$ and $x_3=(132)H$. Then
\begin{align*}
(12)\cdot x_1=x_1,&& (12)\cdot x_2=x_3, && (12)\cdot x_3=x_2,\\
(123)\cdot x_1=x_2, && (123)\cdot x_2=x_3, &&(123)\cdot x_3=x_1.
\end{align*}
Since $G=\langle (12),(123)\rangle$, one has the group
homomorphism 
$\rho\colon G\to\Sym_{X}\simeq\Sym_3$ given by $(12)\mapsto (23)$, $(123)\mapsto (123)$.
\end{example}


\begin{example}
Let $G=Q_8=\{1,-1,i,-i,j,-j,k,-k\}$ and $N=\{1,-1,i,-i\}$. Since
$N$ is normal in $G$, $G$
acts by conjugation on $X=N$.
If $x_1=1$, $x_2=-1$, $x_3=i$ and $x_4=-i$, then
$i\cdot x=x$ for all $x\in N$. Moreover, 
\begin{align*}
j\cdot x_1=x_1, && j\cdot x_2=x_2, && j\cdot x_3=x_4, && j\cdot x_4=x_3.
\end{align*}
Since $G=\langle i,j\rangle$, a group homomorphism $\rho\colon G\to\Sym_X\simeq\Sym_4$ is determined by 
$\rho_i=\id$ and $\rho_j=(34)$.
\end{example}

The following example is important. 

\begin{example}
    The group $\Sym_n$ acts on $\R^n$ by
    \[
    \sigma\cdot (x_1,\dots,x_n)=(x_{\sigma^{-1}(1)},\dots,x_{\sigma^{-1}(n)}).
    \]
    It is very important to use $\sigma^{-1}$ and not 
    $\sigma$, as we need to permute the elements of the standard basis of $\R^3$.

    As a concrete example, let us see that the operation 
    \[
    \sigma\cdot (x_1,x_2,x_3)=(x_{\sigma(1)},x_{\sigma(2)},x_{\sigma(3)})
    \]
    is not an action of $\Sym_3$ on $\R^3$.
    If $\sigma=(12)$ and $\tau=(23)$, then $\sigma\tau=(123)$. Since 
    \begin{align*}
    &(123)\cdot (5,6,7)=(6,7,5),\\
    &(12)\cdot ((23)\cdot (5,6,7))=(1,2)\cdot (5,7,6)=(7,5,6),
    \end{align*}
    this does not define an action. If we compute 
    \begin{align*}
        \sigma\cdot (\tau\cdot (x_1,\dots,x_n))
        =\sigma\cdot (x_{\tau(1)},\dots,x_{\tau(n)})
    \end{align*}
    and for each $i\in\{1,\dots,n\}$ we set $y_i=x_{\tau(i)}$, then 
    \[
    \sigma\cdot (\tau\cdot (x_1,\dots,x_n))=\sigma\cdot (y_1,\dots,y_n)=(y_{\sigma(1)},\dots,y_{\sigma(n)})
    =(x_{\tau\sigma(1)},\dots,x_{\tau\sigma(n)}),
    \]
    even if $\sigma$ and $\tau$ do not commute.

    Now we show that by using inverses, we do have an action. 
    For $j\in\{1,\dots,n\}$, let $y_j=x_{\tau(j)}$,
    that is 
    \[
    (y_1,y_2,\dots,y_n)=\tau\cdot (x_1,x_2,\cdots,x_n)=(x_{\tau^{-1}(1)},x_{\tau^{-1}(2)},\dots,x_{\tau^{-1}(n)}).
    \]
    Then 
    \begin{align*}
        \sigma\cdot (\tau\cdot (x_1,x_2,\dots,x_n))&=\sigma\cdot (y_1,y_2,\dots,y_n)\\
        &=\left(y_{\sigma^{-1}(1)},y_{\sigma^{-1}(2)},\dots,y_{\sigma^{-1}(n)}\right)\\
        &=\left(x_{\tau^{-1}(\sigma^{-1}(1))},x_{\tau^{-1}(\sigma^{-1}(2))},\dots,x_{\tau^{-1}(\sigma
^{-1}(n))}\right)\\
        &=\left(x_{(\sigma\tau)^{-1}(1))},x_{(\sigma\tau)^{-1}(2))},\dots,x_{(\sigma\tau)^{-1}(n))}\right).
    \end{align*}
\end{example}

The following example is also important: 

\begin{example}
    The group $\Sym_n$ acts on the set of polynomials on
    $n$ variables $X_1,\dots,X_n$
    by permitting the variables. For example, for three variables, if 
    $\sigma=(123)$ y $f=X_2X_3-X_1+5X_2X_3^2X_1$, then 
    $\sigma\cdot f=X_2^2X_3-X_1+5X_2X_3^2X_1$.

    Restricting the action, we see that 
    $\Sym_n$ acts on the set 
    \[
    \{\lambda_1X_1+\cdots\lambda_nX_n:\lambda_1,\dots,\lambda_n\in\R\}.
    \]
    Then 
    \begin{align*}
    \sigma \cdot (\lambda_1X_1+\cdots+\lambda_nX_n) &= (\lambda_1X_{\sigma(1)}+\cdots+\lambda_nX_{\sigma(n)})
    =(\lambda_{\sigma(1)}X_1+\cdots+\lambda_{\sigma(n)}X_n).
    \end{align*}
\end{example}

It is relevant to compute the kernel of the action homomorphism. 

\begin{example}
Let $G$ be a group and $H$ be a subgroup of $G$. Then $G$ 
acts on $G/H$ by left multiplication, that is 
$g\cdot (xH)=(gx)H$ for all $g,x\in G$. Let $\rho\colon G\to\Sym_{G/H}$ be the group homomorphism induced by the action. 

We claim that $\ker\rho=\cap_{x\in G}xHx^{-1}$. 
We first prove $\supseteq$. If $g\in xHx^{-1}$ for all 
$x\in G$, then, for a fixed $x\in G$,
 \[
 \rho(g)(xH)=g\cdot (xH)=(gx)H=(xhx^{-1})xH=(xh)H=xH
 \]
because $g=xhx^{-1}$ for some $h\in H$. Thus $\rho(g)=\id$ 
and hence $g\in\ker\rho$. We now prove 
$\subseteq$. If $g\in\ker\rho$, then
 $\rho(g)=\id$. So for all $x\in G$,
 \begin{align*}
\rho(g)(xH)=xH
\Longleftrightarrow (gx)H=xH
\Longleftrightarrow x^{-1}gx\in H
\Longleftrightarrow g\in xHx^{-1}.
 \end{align*}
It is an exercise to show that
$\ker\rho$ is the largest normal subgroup of $G$ 
contained in~$H$.
\end{example}

With these results, we can provide a third 
solution to Exercise~\ref{xca:p_smallest} of 
page~\pageref{xca:p_smallest}.
We let  $G$ act on $G/H$ by left multiplication. 
The induced group homomorphism  $\rho\colon G\to\Sym_p$ has 
kernel 
\[
K=\ker\rho=\bigcap_{x\in G}xHx^{-1}\subseteq H.
\]
By the first isomorphism theorem, 
$G/K\simeq\rho(G)\lesssim\Sym_p$ (this means that 
$\rho(G)$ is isomorphic to a subgroup of $\Sym_p$). 
Thus $|G/K|$ divides $p!$.
Let $m=(H:K)$. By Lagrange's theorem,
\[
(G:K)=(G:H)(H:K)=pm
\]
and hence $pm$ divides $p!$. This implies that $m$ divides $(p-1)!$. If $q$ a prime number dividing 
 $m$, then $q\geq p$, by the minimality of $p$. Moreover, 
 every prime factor of $(p-1)!$ is 
 $<p$. Hence $m=1$ and therefore $H=K$.

\begin{exercise}
    Let a group $G$ act on a set $X$. 
    On $X$, we define the following relation: $x\sim y$ if and only if
    there exists $g\in G$ such that $g\cdot x=y$. Prove 
    that this is an equivalence relation on $X$. 
\end{exercise}

\begin{definition}
\index{Orbit}
Let $G$ be a group acting on a set $X$. If $x\in X$, the
orbit of $x$ is the set
\[
G\cdot x=\{g\cdot x:g\in G\}.
\]
\end{definition}

The orbits of the action of $G$ on $X$ are 
the equivalence classes of the equivalence 
relation induced by the action. In particular, 
every two orbits will be either disjoint or equal. Moreover, 
$X$ decomposes as a disjoint union of orbits. 

\begin{definition}
        \index{Stabilizer of a point}
        Let $G$ be a group that acts on $X$. If $x\in X$, the \textbf{stabilizer} of $x$ in $G$
        is the set   
        \[
        G_x=\{g\in G:g\cdot x=x\}.
        \]
\end{definition}

The reader must prove that the stabilizer is a subgroup. 

\begin{definition}
\index{Action!transitive}
We say that an action of a group $G$ on a set $X$
is \textbf{transitive} if for any $x,y\in X$ there exists $g\in G$ such that $g\cdot x=y$.
\end{definition}

\begin{example}
    Let $G$ be a group and $H$ a subgroup of $G$. Let $G$ act
    on $G/H$ by left multiplication. The action is transitive: if 
    $xH,yH\in G/H$, there exists $g\in G$ such that
    $(gx)H=yH$ (take for example $g=yx^{-1}$). 
\end{example}

\begin{example}
Por evaluación,
el grupo simétrico $\Sym_n$ actúa transitivamente en el conjunto $\{1,\dots,n\}$.
\end{example}

In the definition of a transitive action, there is no assumption
on the number of elements $g$ such that $g\cdot x=y$.

\begin{definition}
\index{Action!faithful}
We say that an action of a group $G$ on a set 
$X$ is \textbf{faithful} if 
\[
\{g\in G:g\cdot x=x\text{ for all $x\in X$}\}=\{1\}.
\]
\end{definition}

The definition is equivalent to the injectivity of 
the group homomorphism induced
by the action. 

\begin{theorem}[Fundamental counting principle]
\index{Theorem!fundamental counting principle}
\label{thm:fundamental}
Let $G$ be a finite group acting on a finite set $X$. If 
$x\in X$, then $|G\cdot x|=(G:G_x)$.
\end{theorem}

\begin{proof}
    Let $\varphi\colon G/G_x\to G\cdot x$, $gG_x\mapsto g\cdot x$. Then $\varphi$ is well-defined, as 
        \[
        gG_x=hG_x\implies h^{-1}g\in G_x
        \implies h^{-1}g\cdot x=x\implies g\cdot x=h\cdot x.
        \]
    Moreover, $\varphi$ is injective: 
        \[
        \varphi(gG_x)=\varphi(hG_x)\implies
        g\cdot x=h\cdot x\implies
        h^{-1}g\in G_x\implies gG_x=hG_x.
        \]
    Finally, $\varphi$ is surjective. Hence 
    $|G/G_x|=|G\cdot x|$.
\end{proof}

Theorem \ref{thm:fundamental} is also known 
as the orbit--stabilizer theorem. 

If $G$ is a group and $X$ and $Y$ are $G$-sets, 
we say that a map $\varphi\colon X\to Y$ is a 
\textbf{homomorphism} of $G$-sets if $\varphi(g\cdot x)=g\cdot \varphi(x)$ for all $g\in G$ y $x\in X$. The bijection 
$\varphi$ constructed in the proof of Theorem \ref{thm:fundamental}
is a homomorphism of $G$-sets, where
$G$ acts on $G/G_x$ by left multiplication: 
\[
\varphi(g\cdot hG_x)=\varphi((gh)G_x)=(gh)\cdot x=g\cdot (h\cdot x)=g\cdot\varphi(hG_x).
\]
Thus $G\cdot x\simeq G/G_x$ as $G$-sets.

\begin{example}
    If$G$ acts on $G$ by conjugation, that is $g\cdot x=gxg^{-1}$, the orbits of this action are called the \textbf{conjugacy classes} 
    of $G$. They are sets of the form
        \[
        G\cdot x=\{gxg^{-1}:g\in G\}.
        \]
    In particular, $G$ decomposes as a disjoint union of conjugacy classes. Moreover, 
    the stabilizers are the centralizers:
        \[
        G_x=\{g\in G:g\cdot x=x\}=\{g\in G:gxg^{-1}=x\}=C_G(x).
        \]
    In particular, $|G\cdot x|=(G:C_G(x))$.
\end{example}

\begin{example}
    Let $H$ be a subgroup of $G$ and $X$ the set of subsets of $G$. Let $G$ act on 
    $X$ by conjugation, that is $S\in X$. Then
        $g\cdot S=gSg^{-1}$. The orbit of $H$ 
        is 
        \[
        G\cdot H=\{g\cdot H:g\in G\}=\{gHg^{-1}:g\in G\},
        \]
        the set of conjugates of $H$. The stabilizer of $H$ in $G$ 
        is 
        \[
        G_H=\{g\in G:g\cdot H=H\}=\{g\in G:gHg^{-1}=H\}=N_G(H),
        \]
        the normalizer of $H$ in $G$. It follows that
        $H$ has exactly $(G:N_G(H))$ conjugates in $G$. In particular,
        if $G$ is finite, 
        the number of conjugates of $H$ divides $|G|$. 
\end{example}

As an application, we provide an alternative proof
of Theorem~\ref{thm:|HK|}. 

\begin{example}
\label{exa:for_HK}
Let $G$ be a group and $H$ and $K$ be subgroups of $G$. 
The group $L=H\times K$ acts on $X=HK$ by 
\[
(h,k)\cdot x=hxk^{-1},\quad x\in X,\,h\in H,\,k\in K.
\]
Note that $1\in HK$ and the action of $L$ on $X$ is transitive, as 
$(h,k^{-1})\cdot 1 = hk$. Since 
\[
L_1=\{(h,k)\in H\times K: (h,k)\cdot 1=1\}=\{(h,k)\in H\times K:h=k\},
\]
it follows that $|L_1|=|H\cap K|$ because there exists a bijection
between $L_1$ and 
$H\cap K$. By the fundamental counting principle, 
\[
|HK|=(L:L_1)=\frac{|H\times K|}{|H\cap K|}=\frac{|H||K|}{|H\cap K|}.
\]
\end{example}

The idea used in Example \ref{exa:for_HK} can be generalized. 

\begin{example}
\index{Double coset}
Let $G$ be a group and $H$ and $K$ be subgroups of $G$. Let the group $L=H\times K$ act on $G$ by
\[
(h,k)\cdot g=hgk^{-1}.
\]
The orbtis are sets of the form 
\[
HgK=\{hgk:h\in H,\,k\in K\}.
\]
These sets are called \textbf{double $(H,K)$-coset}. 
In particular, two double cosets are either disjoint 
or equal. Moreover, $G$ admits a decomposition 
as a disjoint union of double cosets, that is 
\[
G=\bigcup_{i\in I}Hg_iK,
\]
for some set $I$. Now we compute 
\[
L_g=\{(h,k)\in H\times K:hgk^{-1}=g\}=\{(h,g^{-1}hg)\in H\times K\}.
\]
Then $|L_g|=|H\cap gKg^{-1}|$, because there is a bijection between $L_g$ and
$H\cap gKg^{-1}$. By the fundamental counting principle (Theorem~\ref{thm:fundamental}), 
\[
|HgK|=(L:L_g)=\frac{|H\times K|}{|H\cap gKg^{-1}|}=\frac{|H||K|}{|H\cap gKg^{-1}|}.
\]
\end{example}

As another application, we compute the
order of the group $\GL_n(p)$ para $n\geq1$ and 
a prime number $p$. 
The argument also works for
the group $\GL_n(q)$ in the case where
$q$ is a power of the prime number $p$.

\begin{example}
Let $K=\Z/p$.
We claim that 
\[
|\GL_n(p)|=(p^n-1)p^{n-1}|\GL_{n-1}(p)|,
\]
and hence 
\[
|\GL_n(p)|=(p^n-1)(p^n-p)\cdots (p^n-p^{n-1}).
\]
The formula is valid if $n\in\{1,2\}$. 
Assume that it holds for $n-1\geq1$.
The group $G=\GL_{n}(p)$ acts on
$K^{n}$ by left multiplication. There are two orbits, so 
\[
X=\{0\}\cup (K^{n}\setminus\{0\}),
\]
as if $v,w\in K^{n}\setminus\{0\}$, then there exists 
$g\in G$ such that $gv=w$.
By the fundamental counting principle,  
\[
p^{n+1}-1=|K^{n+1}\setminus\{0\}|=(G:G_{e_1}),
\]
where $e_1=(1,0,\dots,0)^T$. If $g=(g_{ij})\in G$ is such that
$ge_1=e_1$, then 
\[
g=
\begin{pmatrix}
1 & g_{12} & \cdots & g_{1n}\\
0 & g_{22} & \cdots & g_{2n}\\
\vdots & \vdots & \ddots &\vdots\\
0 & g_{n1} & \cdots & g_{nn}
\end{pmatrix}.
\]
Therefore $|G_{e_1}|=p^{n-1}|\GL_{n-1}(p)|$, as the submatrix 
$(g_{ij})_{2\leq i,j\leq n}$ is invertible and the 
$g_{1j}$'s can be chosen 
arbitrarily for all $j\in\{2,\dots,n\}$.
Hence 
\[
p^{n}-1=\frac{|G|}{|G_{e_1}|}=\frac{|\GL_n(p)|}{p^{n-1}|\GL_{n-1}(p)|},
\]
which implies the formula we wanted to prove.
\end{example}


